Comments on Blair and Sambanis, 2020, \"Forecasting Civil Wars: Theory and Structure in an Age of 'Big Data' and Machine Learning\", *JCR*

### 2020-06-06 paper.Rmd
getting closer

While replicating and analyzing B&S's results, we found several issues worthy of further discussion and investigation.
These are a) the use of smoothed ROC curves to draw conclusions about which model is best, b) an incorrect implementation of the weighted by ICEWS data model, c)  an inconsistent test set $ for the models examined, d) incorrect scoring of the 2016 forecasts, e) an inappropriate use of random forrest implementation, with very unusual hyperparameters.  Our belief is that these research decisions lead B&S to draw conclusion that are incorrect. The escalation model is not the best and it actually performs worse that the atheoretical, garbage can model with over $1000$ variables.  We turn to discussing these five issues below.

@@

NOW SHOW HOW THIS MATTERS?

@@

GIVE EXAMPLE OF IMPACT HERE?

@@

Give EXAMPLE OF IMPACT HERE?

@@

B&S report confusion matrices for their forecasts for the first half of 2016 (2016-H1) in their Table 4. To create the confusion matrices, B&S treat the 30 highest ranked predictions as positive predictions ("1s") and the rest as negative predictions ("0s"). We replicate their Table 4 in the top portion of \ref{tab:table4}. There are two confusion matrices for slightly different codings of outcomes in 2016-H1, under the corresponding "Assuming Persistence" and "Assuming Change" headings. We should not that the "Assuming Persistence" corresponds to the values in the replication data; the small variations for the "Assuming Change" version are hand-coded in the replication code file that generates the confusion matrices.

We can see that the original table presents 15 or 16 positive cases for 2016-H1, depending on teh DV coding variation. This corresponds to a positive rate of around 9.5% for the first half of 2016 data. In contrast, the corresponding 6-month version of the data from 2001 to 2015, with 30 half-years, has in total 20 civil war onset events, for a much lower positive rate of around 0.5%. Since the positive event rate in the confusion matrices far exceeds the rate of observed civil war onsets in both the training and test data, we suspected that the forecasts were erroneously assessed using civil war incidence, not onset. As we showed above, by examining the replication code we were able to verify that the forecasts were scored using civil war incidence, i.e. including ongoing civil wars, rather than civil war onset years only.

@@

The correct confusion matrices when using observed onset (or the lack of it) are shown in the second part of Table \ref{tab:table4}. In the default "Assuming Persistence" coding, there are no civil war onsets in the data for 2016-H1. Thus the recall values is undefined, while the precision is 0/30 = 0, compared to recall and precision values of 13/15 = 0.87 and 13/30 = 0.43 with the incorrect scoring. The alternative coding ("Assuming Change") produces 2 civil war onsets. Recall is 1 compared to 14/16 = 0.88 before, and precision is 2/30 = 0.07 instead of 14/30 = 0.47.

@@

## AUC-ROC sensitivity

RNG and variation

## Pre-registration

- The escalation model specification, prior to looking at data!
- The hyper-parameter tuning strategy.

@@

B&S have a clear misunderstanding of the role that out-of-sample prediction can plan in analysis.  On the one hand it can be used for simple forecasting while on the other it can be used to evaluate the performance of models, specifically to overfitting and bias.  They further misrepresent cross-validation which is not about fishing for results, as claimed by B&S.  The current state-of-the-art uses prediction and cross-validation most frequently to provide supportive evidence that is independent of the estimation procedure and the in-sample data. It is incorrect to paint this procedure as atheoretical since many studies will have some explanation for how the model was constructed. These procedures may simply be used to provide evidence for a theoretical argument.  Though that argument may be unconvincing and/or atheoretical.

The B&S approach that is advocated is to use theory to guide prediction. But theory is a quite ambiguous and undefined concept.  It is not a procedure.  What they actually do is to create a model with four right hand side variables that is supposed to capture a complicated repression-dissent dynamic.  The dynamic is probably not linear, but their model is. There is a wide-ranging literature on this dynamic that they do not rely on to construct their model.  As such their baseline comparison is hardly a standard bearer for strong theory.

@@

We encountered several issues in the code underlying the B&S analysis.

The issues we encountered are not subjective modeling choices.

When we fix these issue and perform an updated analysis, the conclusions B&S draw all essentially overturn. In other words, B&S findings are based on a faulty analysis, and invalid.

Using the same analysis B&S intend to use, we in fact find that:

- the theory-driven escalation model is outperformed both by the low-effort 1,160 predictor all CAMEO model and the Average ensemble model
- structural variables substantially improve the escalation model's performance when added to the pool of predictors on which the underlying random forest model is based.

>>>>

While replicating and analyzing B&S's results, we found several issues worthy of further discussion and investigation.
These are a) the use of smoothed ROC curves to draw conclusions about which model is best, b) an incorrect implementation of the weighted by ICEWS data model, c)  an inconsistent test set $ for the models examined, d) incorrect scoring of the 2016 forecasts, e) an inappropriate use of random forrest implementation, with very unusual hyperparameters.  Our belief is that these research decisions lead B&S to draw conclusion that are incorrect. The escalation model is not the best and it actually performs worse that the atheoretical, garbage can model with over $1000$ variables.  We turn to discussing these five issues below.  We hold a complete analysis that corrects all these issues until later, as there are many possible permuations of a seriatim unfolding.

@@

@@

@@

@@

B&S report confusion matrices for their forecasts for the first half of 2016 (2016-H1) in their Table 4. To create the confusion matrices, B&S treat the 30 highest ranked predictions as positive predictions ("1s") and the rest as negative predictions ("0s"). We replicate their Table 4 in the top portion of \ref{tab:table4}. There are two confusion matrices for slightly different codings of outcomes in 2016-H1, under the corresponding "Assuming Persistence" and "Assuming Change" headings. We should note that the "Assuming Persistence" corresponds to the values in the replication data; the small variations for the "Assuming Change" version are hand-coded in the replication code file that generates the confusion matrices and appear to be subjective assessments of B&S.

We can see that the original table presents 15 or 16 positive cases for 2016-H1, depending on the dependent variable coding variation. This corresponds to a positive rate of around 9.5% for the first half of 2016 data. In contrast, the corresponding 6-month version of the data from 2001 to 2015, with 30 half-years, has in total 20 civil war onset events, for a much lower positive rate of around 0.5%. Since the positive event rate in the confusion matrices far exceeds the rate of observed civil war onsets in both the training and test data, this suggests that the forecasts were erroneously assessed using civil war incidence, not onset. As we showed above, by examining the replication code we were able to verify that the forecasts were scored using civil war incidence, i.e. including ongoing civil wars, rather than civil war onset years only.

@@

The correct confusion matrices when using observed onset (or the lack of it) are shown in the second part of Table \ref{tab:table4}. In the default "Assuming Persistence" coding, there are no civil war onsets in the data for 2016-H1. Thus, the recall values is undefined, while the precision is 0/30 = 0, compared to reported recall and precision values of 13/15 = 0.87 and 13/30 = 0.43. The alternative coding ("Assuming Change") produces 2 civil war onsets. Recall is 1 compared to 14/16 = 0.88 before, and precision is 2/30 = 0.07 instead of 14/30 = 0.47.

@@

@@

B&S have a clear misunderstanding of the role that out-of-sample prediction can play in analysis.  On the one hand it can be used for simple forecasting, while on the other it can be used to evaluate the performance of models, specifically to overfitting and bias.  They further misrepresent cross-validation---which is not about fishing for results, as claimed by B&S.  The current state-of-the-art uses prediction and cross-validation most frequently to provide supportive evidence that is independent of the estimation procedure and the in-sample data. It is incorrect to paint this procedure as atheoretical since many studies will have some explanation for how the model was constructed. These procedures may simply be used to provide evidence for a theoretical argument.  Though that argument may be unconvincing and/or atheoretical.

The B&S approach that is advocated is to use theory to guide prediction. But theory is an ambiguous and undefined concept.  It is not a procedure.  What they actually do is to create a model with four right hand side variables that is supposed to capture a complicated repression-dissent dynamic.  The dynamic is probably not linear, but their model is. There is a wide-ranging literature on this dynamic that they do not rely on to construct their model.  As such their baseline comparison is a poor standard bearer for strong theory.

@@

We encountered several issues in the code underlying the B&S analysis. The issues we encountered are not subjective modeling choices. When we fix these issues and perform an updated analysis, the B&S conclusions are all essentially overturn. In other words, B&S findings are based on a faulty analysis, and invalid.

Using the same analysis B&S intend to use, we in fact find that:
- the theory-driven escalation model is outperformed both by the low-effort 1,160 predictor all CAMEO model and the Average ensemble model
- structural variables substantially improve the escalation model's performance when added to the pool of predictors on which the underlying random forest model is based.

Blair & Sambanis have focused attention on comparing forecasting models and forecasts in the civil war domain.  As social science becomes more adept at predictive analysis this will doubtless be of increasing importance.  However, these comparisons must be done carefully to ensure that correct inferences are drawn.  We continue to think that theory is overrated (@ward:2016), and that machine learning and big data will allow us to learn new things, but it will be interesting to see how the evidence for these claims are adjudicated with additional usage and careful evaluation.
